---
title: "Forecasting Ontario Electricity Prices using ARMA"
author: "Abdel Salam Abuawad"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

The data is sourced from [IESO Yearly Hourly HOEP Report](https://reports-public.ieso.ca/public/PriceHOEPPredispOR/). Note the prices are in terms of CAD/MWh.
Firstly, we establish the base predictions done by IESO we need to beat.
Secondly, we show that ARMA is sufficient to beat IESO's forcasts.
Finally, we show that the residuals closely simulate a white noise process.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
set.seed(123)
library(stats)
library(forecast)
library(tseries)
library(moments)
library(e1071)
library(knitr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(dplyr)
library(rugarch)
```

## EDA

```{r}
# importing and cleaning the data
IESO25H <- read.csv("data/PUB_PriceHOEPPredispOR_2025_v82.csv") 
colnames(IESO25H) <- as.character(IESO25H[3,])
IESO25H <- IESO25H[-c(1, 2,3), ]
IESO25H <- IESO25H %>%
mutate(HOEP = as.numeric(HOEP),
         H1_predicted = as.numeric(`Hour 1 Predispatch`),
         H2_predicted = as.numeric(`Hour 2 Predispatch`),
         H3_predicted = as.numeric(`Hour 3 Predispatch`))
IESO25H$Timestamp <- as.POSIXct(paste(IESO25H$Date, IESO25H$Hour), format="%Y-%m-%d %H")
IESO25H <- IESO25H[, -c(1,2,4,5,6,7,8,9)]
IESO25H <- IESO25H[, c(5,1,2,3,4)]

# Replace empty values with the median
IESO25H[-1] <- lapply(IESO25H[-1], function(col) ifelse(is.na(col), median(col, na.rm = TRUE), col))

# Hour i Predictions
IESO25H1 <- data.frame(Timestamp = head(IESO25H$Timestamp, -1), HOEP = head(IESO25H$HOEP, -1), H1_predicted = tail(IESO25H$H1_predicted, -1))
IESO25H2 <- data.frame(Timestamp = head(IESO25H$Timestamp, -2), HOEP = head(IESO25H$HOEP, -2), H2_predicted = tail(IESO25H$H2_predicted, -2))
IESO25H3 <- data.frame(Timestamp = head(IESO25H$Timestamp, -3), HOEP = head(IESO25H$HOEP, -3), H3_predicted = tail(IESO25H$H3_predicted, -3))

# Hist

ggplot(IESO25H3, aes(x = HOEP)) + 
  geom_histogram(binwidth = 2, fill = "blue", color = "black") + 
  ggtitle("Electricity Prices Ontario") + 
  xlab("Price") + 
  ylab("Hours")

# TS
ggplot(IESO25H1, aes(x = Timestamp)) +
  geom_line(aes(y = HOEP)) +
  labs(title = "TS plot",
       x = "Timestamp",
       y = "HOEP (CAD)") +
  theme_minimal()


# First four moments
mean_hoep <- mean(IESO25H$HOEP, na.rm = TRUE)
variance_hoep <- var(IESO25H$HOEP, na.rm = TRUE)
skewness_hoep <- skewness(IESO25H$HOEP)
kurtosis_hoep <- kurtosis(IESO25H$HOEP, na.rm = TRUE)


# Regular Moments
moments_table <- data.frame(
  Moment = c("Mean", "Variance", "Skewness", "Kurtosis"),
  Value = c(mean_hoep, variance_hoep, skewness_hoep, kurtosis_hoep)
)

# Display the table in a clean format
kable(moments_table, col.names = c("Moment", "Value"), caption = "First four moments")
```
## Filtering out outliers 3 std. dev. away

```{r}

# 4 Moments after Filtering Outliers
mean_value <- mean(IESO25H$HOEP, na.rm = TRUE)  # Calculate the mean
std_dev <- sd(IESO25H$HOEP, na.rm = TRUE)      # Calculate the standard deviation

# Filter data to exclude points more than 3 standard deviations from the mean
filtered <- IESO25H[abs(IESO25H$HOEP - mean_value) <= 3 * std_dev, ]

outlier_count <- length(IESO25H[abs(IESO25H$HOEP - mean_value) > 3 * std_dev, ]$HOEP) 
total_count <- length(IESO25H$HOEP)
percent_outliers <- length(IESO25H[abs(IESO25H$HOEP - mean_value) > 3 * std_dev, ]$HOEP) / length(IESO25H$HOEP) 
print(paste("Outlier count:", outlier_count))
print(paste("Total count:", total_count))
print(paste("Percent of outliers:", percent_outliers))
       
# Filtered Hour i Predictions
IESO25H1 <- data.frame(Timestamp = head(filtered$Timestamp, -1), HOEP = head(filtered$HOEP, -1), H1_predicted = tail(filtered$H1_predicted, -1))
IESO25H2 <- data.frame(Timestamp = head(filtered$Timestamp, -2), HOEP = head(filtered$HOEP, -2), H2_predicted = tail(filtered$H2_predicted, -2))
IESO25H3 <- data.frame(Timestamp = head(filtered$Timestamp, -3), HOEP = head(filtered$HOEP, -3), H3_predicted = tail(filtered$H3_predicted, -3))

# 'filtered' now contains the cleaned data
filtered_kurt <- kurtosis(filtered$HOEP)
filtered_skew <-skewness(filtered$HOEP)
filtered_mean <-mean(filtered$HOEP)
filtered_var <-var(filtered$HOEP)

# Hist

ggplot(filtered, aes(x = HOEP)) + 
  geom_histogram(binwidth = 2, fill = "blue", color = "black") + 
  ggtitle("Electricity Prices Ontario") + 
  xlab("Price") + 
  ylab("Hours")

# TS
ggplot(filtered, aes(x = Timestamp)) +
  geom_line(aes(y = HOEP)) +
  labs(title = "Outliers-Excluded TS plot",
       x = "Timestamp",
       y = "HOEP (CAD)") +
  theme_minimal()



# Filtered Moments
filtered_moments_table <- data.frame(
  Moment = c("Mean", "Variance", "Skewness", "Kurtosis"),
  Value = c(filtered_mean, filtered_var, filtered_skew, filtered_kurt)
)

# Display the table in a clean format
kable(filtered_moments_table, col.names = c("Moment", "Value"), caption = "Excluding Outliers (3 std.)")
```
Motivation for ARMA: the data is stationary according to an ADF test, and an ACF plot suggests some strong autocorrelation across multiple lags. Moreover, the ACF plot exhibits fluctuations, such as a decrease between lags 1 to 5, followed by an increase from lags 5 to 10. This suggests there may be underlying shocks that persist across the data. Thus, we have both autocorrelation, and possibly persistent shocks.
```{r}
adf.test(filtered$HOEP, alternative = "stationary")
acf(filtered$HOEP, main = "ACF Plot")

```
## IESO's Forecasts

```{r}

# first week predictions
IESO25H1Jan <- IESO25H1 %>%
  filter(Timestamp >= as.POSIXct("2025-03-17 01:00:00") & Timestamp <= as.POSIXct("2025-03-24 01:00:00"))
IESO25H2Jan <- IESO25H2 %>%
  filter(Timestamp >= as.POSIXct("2025-03-17 01:00:00") & Timestamp <= as.POSIXct("2025-03-24 01:00:00"))
IESO25H3Jan <- IESO25H3 %>%
  filter(Timestamp >= as.POSIXct("2025-03-17 01:00:00") & Timestamp <= as.POSIXct("2025-03-24 01:00:00"))

filteredH1 <- filtered %>%
  filter(Timestamp >= as.POSIXct("2025-03-17 01:00:00") & Timestamp <= as.POSIXct("2025-03-24 01:00:00"))



# Plot actual HOEP vs predicted HOEP over time
ggplot(IESO25H1Jan, aes(x = Timestamp)) +
  geom_line(aes(y = HOEP, color = "Actual")) +
  geom_line(aes(y = H1_predicted, color = "Predicted")) +
  labs(title = "Actual vs Predicted HOEP 1 step",
       x = "Timestamp",
       y = "HOEP (CAD)",
       color = "Legend") +
  theme_minimal()

ggplot(IESO25H2Jan, aes(x = Timestamp)) +
  geom_line(aes(y = HOEP, color = "Actual")) +
  geom_line(aes(y = H2_predicted, color = "Predicted")) +
  labs(title = "Actual vs Predicted HOEP 2 steps",
       x = "Timestamp",
       y = "HOEP (CAD)",
       color = "Legend") +
  theme_minimal()

ggplot(IESO25H3Jan, aes(x = Timestamp)) +
  geom_line(aes(y = HOEP, color = "Actual")) +
  geom_line(aes(y = H3_predicted, color = "Predicted")) +
  labs(title = "Actual vs Predicted HOEP 3 steps",
       x = "Timestamp",
       y = "HOEP (CAD)",
       color = "Legend") +
  theme_minimal()

# Residuals for 1 step
residuals_ieso <- data.frame(residuals = IESO25H1Jan$HOEP - IESO25H1Jan$H1_predicted,
                               Timestamp = IESO25H1Jan$Timestamp)
ggplot(residuals_ieso, aes(x = Timestamp)) +
  geom_line(aes(y = residuals, color = "Actual")) +
  labs(title = "Residuals of IESO 1 Step",
       x = "Timestamp",
       color = "Legend") +
  theme_minimal()

Box.test(residuals_ieso$residuals, lag = 25, type = "Ljung-Box")
acf(residuals_ieso$residuals, main = "ACF of Residuals")
Box.test(residuals_ieso$residuals, lag = 25,  type = "Box-Pierce")
adf.test(residuals_ieso$residuals)
```

## ARMA Forecasts

```{r}
hours <- 24*7

if (nrow(filtered) > hours) {
filtered_forecasts <- c()

# Loop from 24 down to 1
for (i in hours:1) {
  
  # Subset the 'filtered' data to exclude the last 'i' rows for training
  filtered_subset <- filtered$HOEP[1:(nrow(filtered) - i)]  
  ts_filtered <- ts(filtered_subset)  

  filtered_arma_model <- auto.arima(ts_filtered, d = 0) 
  
  filtered_fc_arma <- forecast(filtered_arma_model, h = 3, stepwise=FALSE, approximation=FALSE)  

  # Forecast
  filtered_forecasts <- c(filtered_forecasts, c(filtered_fc_arma$mean[1]))

}
} else {
  cat("Error \n")
}



results <- data.frame(forecasted = filtered_forecasts,
                      actual = tail(filtered$HOEP, hours),
                      IESOPred1 = tail(filtered$H1_predicted, hours),
                      IESOPred2 = tail(filtered$H2_predicted, hours),
                      IESOPred3 = tail(filtered$H3_predicted, hours),
                      Timestamp = tail(filtered$Timestamp,hours),
                      Residuals = tail(filtered$HOEP, hours) - filtered_forecasts)
filtered_arma_model

```
## Graphical Results of forecasts

```{r}

# Print forecast details.
ggplot(results, aes(x = Timestamp)) +
  geom_line(aes(y = actual, color = "actual")) +
  geom_line(aes(y = forecasted, color = "forecasted")) +
  labs(title = "Actual vs ARMA(3,5) Forecasted",
       x = "Timestamp",
       y = "HOEP (CAD)",
       color = "Legend") +
  theme_minimal()

ggplot(results, aes(x = Timestamp)) +
  geom_line(aes(y = actual, color = "actual")) +
  geom_line(aes(y = IESOPred3, color = "forecasted")) +
  labs(title = "Actual vs IESO 3 step ",
       x = "Timestamp",
       y = "HOEP (CAD)",
       color = "Legend") +
  theme_minimal()


# Forecasted MAE and RMSE's 
arma_mae <- mean(abs(results$actual - results$forecasted))
arma_rmse <- sqrt(mean((results$actual - results$forecasted)^2))

# IESO MAE and RMSE's 
actual_filtered <- results$actual
filtered_1_step <- results$IESOPred1
filtered_2_step <- results$IESOPred2
filtered_3_step <- results$IESOPred3

mae_ieso_1_step <- mean(abs(actual_filtered - filtered_1_step))
mae_ieso_2_step <- mean(abs(actual_filtered - filtered_2_step))
mae_ieso_3_step <- mean(abs(actual_filtered - filtered_3_step))

rmse_ieso_1_step <- sqrt(mean((actual_filtered - filtered_1_step)^2))
rmse_ieso_2_step <- sqrt(mean((actual_filtered - filtered_2_step)^2))
rmse_ieso_3_step <- sqrt(mean((actual_filtered - filtered_3_step)^2))

# Table Comparing IESO's performance with ARMA(3,5)
metrics_table <- data.frame(
  Model = c("IESO 1", "IESO 2", "IESO 3", "ARMA"),
  MAE = c(mae_ieso_1_step, mae_ieso_2_step, mae_ieso_3_step, arma_mae),
  RMSE = c(rmse_ieso_1_step, rmse_ieso_2_step, rmse_ieso_3_step, arma_rmse)
)

kable(metrics_table, col.names = c("Model", "MAE", "RMSE"), caption = "IESO Predictions vs ARMA(3,5)")


```
## Analyzing Residuals

The Ljung-Box and Box-Pierce test both yield large p-values. Furthermore, we have that acf plot shows little autocorrelation across the residuals, and our adf test shows its stationary. Hence, the residuals show behavior of white noise.
```{r}

ggplot(results, aes(x = Timestamp)) +
  geom_line(aes(y = Residuals, color = "Residuals")) +
  labs(title = "Residuals",
       x = "Timestamp",
       y = "HOEP (CAD)",
       color = "Legend") +
  theme_minimal()
Box.test(results$Residuals, lag = 25, type = "Ljung-Box")
acf(results$Residuals, main = "ACF of Residuals")
Box.test(results$Residuals, lag = 25,  type = "Box-Pierce")
adf.test(results$Residuals)



```
## Comparing Moments

```{r}

# Compute the first four moments for results$actual
mean_actual <- mean(results$actual, na.rm = TRUE)
var_actual <- var(results$actual, na.rm = TRUE)
skew_actual <- skewness(results$actual, na.rm = TRUE)
kurt_actual <- kurtosis(results$actual, na.rm = TRUE)

# Compute the first four moments for results$forecasted
mean_forecasted <- mean(results$forecasted, na.rm = TRUE)
var_forecasted <- var(results$forecasted, na.rm = TRUE)
skew_forecasted <- skewness(results$forecasted, na.rm = TRUE)
kurt_forecasted <- kurtosis(results$forecasted, na.rm = TRUE)

# Compute the first four moments for results$IESOPred
mean_IESOPred <- mean(results$IESOPred1, na.rm = TRUE)
var_IESOPred <- var(results$IESOPred1, na.rm = TRUE)
skew_IESOPred <- skewness(results$IESOPred1, na.rm = TRUE)
kurt_IESOPred <- kurtosis(results$IESOPred1, na.rm = TRUE)

# Compute Euclidean distances
euclidean_actual_forecasted <- sqrt(sum((results$actual - results$forecasted)^2, na.rm = TRUE))
euclidean_actual_IESOPred <- sqrt(sum((results$actual - results$IESOPred1)^2, na.rm = TRUE))

# Create a data frame for the results
comparison_table <- data.frame(
  Metric = c("Mean", "Variance", "Skewness", "Kurtosis", "Euclidean Distance"),
  Actual = c(mean_actual, var_actual, skew_actual, kurt_actual, NA),
  Forecasted = c(mean_forecasted, var_forecasted, skew_forecasted, kurt_forecasted, euclidean_actual_forecasted),
  IESOPred = c(mean_IESOPred, var_IESOPred, skew_IESOPred, kurt_IESOPred, euclidean_actual_IESOPred)
)

# Display the table
kable(comparison_table, col.names = c("Metric", "Actual", "ARMA", "IESOPred"), caption = "Comparison of Moments and Euclidean Distances")
```





